\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-03 paper/paper.tex
paper updates on friday

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her prior personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''

The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

@@

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit
contrived---smiling like ``Honigkuchenpferde,'' cookies in the shape
of a horse---and the sports director of the German Athletics
Federation, Thomas Kurschilgen, stirred up controversy when he
suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we touch on this in the analysis that
follows.  The marathon also featured twins from North Korea, Kim
Hye-song and Kim Hye-gyong, who posted identical times and finished
10th and 11th in the race, respectively. The Kim finish, unlike the
Hahner finish, appears devoid of post-race controversy. Moreover,
three triplets from Estonia competed in the Rio marathon, although
only two, Lily Luik and Leila Luik, finished it, in 97th and 114th
place, respectively. The third Estonia triplet, Liina Luik, recorded
what is known as a DNF---an abbreviation that means did not finish, a
term that we will use throughout this article.

@@

The Kurschilgen accusation against the Hahner twins has two
components, that the twins finished simultaneously, holding hands and
smiling, \emph{and} that they ran slowly.  We suspect that Kurschilgen
would not have expressed ire at the Hahners had they finished in 1st
and 2nd place in Rio, hand-in-hand with wide grins, but of course we
do not know this.  As such, our investigation of the charges that
Kurschilgen offered will distinguish between the idea of slow finish
versus a simultaneous finish.

Our research design has two components.  In the first, we present
visualizations that describe various features of the Hahner twin's
results, and an important element of our visualizations is the
difference between a runner's Rio time and her prior personal best
time in the marathon.  Overall, our visualizations suggest blah blah
bhah.  We then turn to modeling and in particular a regression-based
simulation.  In simulated Olympic marathons, which we develop based on
relationships (excluding the Hahner and Kim twins and Luik triplets)
between known personal best times and observed Rio finishing times, we
find that the Hahner twins finished suspiciously close to each other
given the disparty between their personal best times.  OR THEY DID
NOT.  Finally, we return to Kurschilgen claims  about the Hahners and
offer our thoughts about their validity.

@@

Figure \ref{fig:scatter} contains two plots, both of which describe
how the Rio Olympic women's marathon results varied as a function of
athletes' personal best times and their half marathon split times.
The points in each plot are colored by twin/triplet status, and both
plots contain second-order polynomial regression lines.  We treat an
athlete's prior personal best marathon time as a measure of the
athlete's underlying marathon talent, and we interpret an athlete's
half marathon split as a measure of the athlete's skill at marathon
running on August 14, 2016, the date of the women's marathon in Rio.

@@

Considering first Figure \ref{fig:45degreeplot_half}, note that this
figure includes a solid 45-degree line as the horizontal and vertical
axes in the figure correspond to full marathons.  The rug marks on the
bottom of the figure denote the personal best times of runners who
DNFed.  Given the paucity of points (five of them) below the 45-degree
line, the vast majority of Olympic marathoners ran slower in Rio
compared to their personal bests.  Relative to personal best times,
the Hahner twins were definitely on the slow side, but a number of
runners had greater differences between their Olympic times and their
personal bests than the Hahners.  These runners are denoted with
squares in Figure \ref{fig:45degreeplot}, and there are 13 such
symbols, highlighting approximately 10\% of the racers who completed
the marathon.  The figure shows moreover that there were two women who
had personal best times slightly faster than the Hahner's and yet
finished after both of the German women.  Although Figure
\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
one would have expected given their previous best marathon times, it
is not consistent with the accusation that they dramatically slowed
down in the Rio marathon.

@@

One limitation of using personal best times as indicators of
underlying athletic talent is that these times are potentially
confounded by the marathon courses at which they were set (some*sh
courses, like the Berlin marathon, are known for fast times) and race
conditions like weather.  In addition, personal best times may not
capture raceday idiosyncrasies that might affect individual runners.
With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
finishing times against half marathon splots.  This figure has a
regression line as before but no 45-degree line.

Figure \ref{fig:45degreeplot_half} shows that there was nothing
abnormal about the Hahner twins' overall finishing times, conditional
on their half marathon split times.  This is consistent with the
previous Figure \ref{fig:45degreeplot} and inconsistent with the
accusations against the Hahner twins, at least the part of the
accusation that focuses on their excessively slowing down.

@@

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners or was it
unusual?  We might ask, are there other pairs of runners who had a
difference between personal best times of 115 seconds apart and, if
so, how close where their finishing times?

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times in seconds are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
of runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For the 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points about the Hahner's orange dot,
and this shows that, conditional on a 115 second difference in
personal best times, most marathoners did not have close finishing
times like the Hahnes.  Some pairs of runners with 115 second personal
best differences had finishing time differences of 1000 seconds, i.e.,
in excess of 15 minutes.  Of course the points in Figure
\ref{fig:diffdiffscatter} are not independent, but they provide a
sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

Thinking about the accusations leveled against the Hahner twins,
Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
did indeed run with an eye on each other.  In fact, the same can be
said of the Kim twins, who ran seemingly in lockstep throughout the
entire Rio marathon.  The North Korean twins had a personal best
difference of 53 seconds and a finishing time difference of literally
zero seconds.  Beyond these twins, there were eight pairs of runners
with a 53 second personal best difference, and resulting finishing
time differences are as follows: 9, 51, 228, 340, 352, 571, 662, and
751.  As in the Hahner case, the Kim twins compressed their finishing
times more than any other pair of runners with a similar personal best
difference.

To get a sense of the lockstep nature of the Kim twins race, Figure
\ref{fig:secondsbehind} describes each runner's status at the various
split times on the marathon course.  Each dot in the figure depicts a
recorded split and the number of seconds each runner was behind the
race leader at the time.  There are more dots at earlier splits due to
the accumulation of DNFs.  The usual color scheme applies here, and
the Estonian DNF prior to the half marathon split is evident.  The
orange dots in the figure represent the Hahner twins, and as noted in
the figure Lisa Hahner was ahead of her sister through 15 kilometers
at which point Anna surged ahead.  Figure \ref{fig:secondsbehind}
contains two red dots representing the North Korean Kim twins, but
this is not visually apparent because the Kim twins had identical
split times during the entire race marathon.

>>>>

At 9:30am on August 14, 2016, the Women's Olympic marathon kicked off
in Rio de Janeiro, Brazil, when 156 runners from 80 countries across
the world left the starting line en route to their destination 42.195
kilometers away. Two hours, twenty-four minutes, and four seconds
later, Jemima Sumgong of Kenya would be the first to cross the
finishline and take home gold; Sumgong was just three and one-half
minutes slower than her prior personal best time in the
marathon. Approximately 21 minutes later, twin marathoners from
Germany, Anna and Lisa Hahner, would cross the finishline together,
holding hands and celebrating a personal victory. Although the Hahners
would finish 81st and 82nd, respectively, well behind the winners of
the marathon, Anna Hahner would describe their joint finish as a
``magical moment.''

@@

The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. While
many believed the moment was a reflection of the Olympic spirit, not
everyone agreed with this rosy interpretation. The twins' happy facial
expressions at the finish were portrayed as a bit contrived---smiling
like ``Honigkuchenpferde,'' cookies in the shape of a horse---and the
sports director of the German Athletics Federation, Thomas
Kurschilgen, stirred up controversy when he suggested that the
Hahners' photo-finish was no coincidence.  Kurschilgen averred that
the twins slowed down so as to finish simultaneously and create a
spectacle which would ``generate media attention.'' Kurschilgen
justified his charge with the fact that the twins ran in the Rio
marathon at least 18 minutes slower their personal best times prior to
the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties.  The marathon also featured twins from North
Korea, Kim Hye-song and Kim Hye-gyong, who posted identical times and
finished 10th and 11th in the race, respectively. The Kim finish,
unlike the Hahner finish, appears devoid of post-race
controversy. Moreover, three triplets from Estonia competed in the Rio
marathon, although only two, Lily Luik and Leila Luik, finished it, in
97th and 114th place, respectively. The third Estonia triplet, Liina
Luik, recorded what is known as a DNF---an abbreviation that means did
not finish.  Although our focus here is the Hahner twins, we touch on
the Kim twins and Luik triplets in the course of what follows.

@@

The Kurschilgen accusation against the Hahner twins has two
components, that these two women ran slowly \emph{and} that they
finished simultaneously.  We suspect that Kurschilgen would not have
expressed ire at the Hahners had they finished in 1st and 2nd place in
Rio, hand-in-hand with wide grins, but of course we do not know this.
As such, our investigation of the charges that Kurschilgen offered
will distinguish between the idea of slow finish versus a simultaneous
finish.

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between a
runner's Rio time and her prior personal best time in the marathon.
Overall, our visualizations suggest that the Hahner twins' pace in the
marathon was slow, but not excessively so, but that their simultaneous
finish was quite unusual given the twins' differences in abilities
(and similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, which we develop based on relationships
(excluding the Hahner and Kim twins and Luik triplets) between known
personal best times and observed Rio finishing times.  Our simulations
show that the Hahner twins finished suspiciously close to each other
given the disparty between their personal best times.  Finally, we
return to Kurschilgen claims about the Hahners and offer our thoughts
about their validity.

@@

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio Olympic women's marathon varied as a
function of athletes' personal best times and their Rio half marathon
split times.  The points in the plots are colored by twin/triplet
status, and both plots contain second-order polynomial regression
lines.  We treat an athlete's prior personal best marathon time as a
measure of the athlete's underlying marathon talent, and we interpret
an athlete's half marathon split as a measure of the athlete's skill
at marathon running on August 14, 2016, the date of the women's
marathon in Rio.

@@

Figure \ref{fig:45degreeplot_half} includes a solid 45-degree line as
the axes in the figure correspond to full marathons; in addition, the
figure's rug marks denote personal best times of runners who DNFed.
Given the paucity of points (five of them) below the pictured
45-degree line, the vast majority of Olympic marathoners ran slower in
Rio compared to their personal bests.  Relative to personal best
times, the Hahner twins were definitely on the slow side, but a number
of runners had greater differences between their Olympic times and
their personal bests than the Hahners.  These runners are denoted with
squares in Figure \ref{fig:45degreeplot}, and there are 13 such
symbols, highlighting approximately 10\% of the racers who completed
the marathon.  The figure shows moreover that there were two women who
had personal best times slightly faster than the Hahner's and yet
finished after the German twins.  Although Figure
\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
one would have expected given their previous best marathon times, it
is not consistent with the accusation that they dramatically slowed
down in the Rio marathon.

@@

One limitation of using personal best times as indicators of
underlying athletic talent is that these times are potentially
confounded by the marathon courses at which they were set (some
courses, like the Berlin marathon, are known for fast times) and race
conditions like weather.  In addition, personal best times may not
capture raceday idiosyncrasies that might affect individual runners.
With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
finishing times against half marathon splits.  This figure has a
regression line as before but no 45-degree line.

Figure \ref{fig:45degreeplot_half} shows that there was nothing
abnormal about the Hahner twins' overall finishing times, conditional
on their half marathon split times.  This is consistent with the
previous Figure \ref{fig:45degreeplot} and inconsistent with
accusations made against the Hahner twins, at least the part of the
accusation that focused on their  overall pace.

@@

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners or, in
contrast, was it unusual?  Are there other pairs of marathoners who
had a difference between personal best times of 115 seconds apart and,
if so, how close where their finishing times?

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times, in seconds, are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
of runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points about the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahnes.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

Thinking about the accusations leveled against the Hahner twins,
Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
did indeed run with an eye on each other.  In fact, the same can be
said of the Kim twins, who ran seemingly in lockstep throughout the
entire Rio marathon.  The North Korean twins had a personal best
difference of 53 seconds and a finishing time difference of literally
zero seconds.  Beyond these twins, there were eight pairs of Rio
runners with a 53 second personal best difference, and resulting
finishing time differences are as follows: 9, 51, 228, 340, 352, 571,
662, and 751.  As in the Hahner case, the Kim twins compressed their
finishing times more than any other pair of runners with a similar
personal best difference.

To get a sense of the lockstep nature of the Kim twins race, Figure
\ref{fig:secondsbehind} describes each runner's status at the various
split times on the marathon course.  Each dot in the figure depicts a
recorded split and the number of seconds each runner was behind the
race leader at the time.  There are more dots at earlier splits due to
the accumulation of DNFs.  The usual color scheme applies here, and
the Estonian DNF prior to the half marathon split is evident.  The
orange dots in the figure represent the Hahner twins, and as noted in
the figure Lisa Hahner was ahead of her sister through 15 kilometers
at which point Anna surged.  Figure \ref{fig:secondsbehind} contains
two red dots representing the North Korean Kim twins, but this is not
visually apparent because the Kim twins had identical split times
during the entire race marathon.
